---
title: Convergence of SMACOF
author:
  - name: Jan de Leeuw
    orcid: 0000-0003-1420-1797
    email: jan@deleeuwpdx.net
    affiliation:
      - name: University of California Los Angeles
        city: Los Angeles
        state: CA
        url: www.ucla.edu
    license: "CC0"
date: last-modified
date-format: long
bibliography: [mypubs.bib, total.bib]
number-sections: true
pdf-engine: lualatex
keep-tex: true
format:
  pdf:
    fontsize: 12pt
    include-in-header:
      - preamble.tex
    keep-tex: true
    link-citations: true
    documentclass: scrartcl
    number-sections: true
  html:
    fontsize: 12pt
    include-in-header:
      - preamble.css
    keep-md: true
    number-sections: true
toc: true
toc-depth: 3
editor: source
papersize: letter
graphics: true
link-citations: true
mainfont: Times New Roman
abstract: TBD
---

```{r}
#| label: loadpackages
#| echo: false

suppressPackageStartupMessages(library(knitr, quietly = TRUE))
suppressPackageStartupMessages(library(tinytex, quietly = TRUE))
suppressPackageStartupMessages(library(numDeriv, quietly = TRUE))
```

```{r}
#| label: function_code
#| echo: false

source("smacof.R")
source("derivative.R")
```
```{r}
#| label: packages
#| echo: false

options(digits = 10) 
```

# Introduction

The components of a (Euclidean, metric, least squares, r-dimensional) multidimensional scaling (MDS) 
problem are

* a symmetric, non-negative, and
hollow\footnote{A hollow matrix has a zero diagonal.} matrix $\Delta$ of *dissimilarities*,
* a symmetric, non-negative, and hollow matrix $W$ of *weights*,
* a list of $m$ *configurations* $Y_1,\cdots,Y_m$ of dimension $n\times r$.



we minimize
the real valued loss function\footnote{The symbol $:=$ is used for definitions.}
\begin{equation}
\sigma(X):=\frac12\mathop{\sum}_{1\leq i<j\leq n} w_{ij}(\delta_{ij}-d_{ij}(X))^2,
\label{eq-stressx}
\end{equation}
defined on $\mathbb{R}^{n\times p}$, the space of all $n\times p$ matrices. We
call $\sigma(X)$ the *stress* of *configuration* $X$. Minimizing stress over all
$p-$dimensional configurations (over *configuration space*) is a classical metric
MDS problem (@kruskal_64a, @kruskal_64b). The parameters of the problem are the
$np$ elements of $X$.

In \eqref{eq-stressx} the matrices of *weights* $W=\{w_{ij}\}$ and
*dissimilarities* $\Delta=\{\delta_{ij}\}$ are . The matrix-valued
function $D(X)=\{d_{ij}(X)\}$ contains Euclidean distances between the rows of
the configuration $X$, which are the coordinates of $n$ *points* in
$\mathbb{R}^p$. Thus $D(X)$ is also symmetric, non-negative, and hollow.

We also assume, without any real loss of generality, that $W$ is *irreducible*,
so that the MDS problem does not separate into a number of smaller MDS problems
(@deleeuw_C_77). For the symmetric matrix $W$ irreducibility means that we
cannot find a permutation of rows and columns that make the permuted $W$
block-diagonal, i.e. the direct
sum\footnote{The direct sum of a number of matrices $X_1,\dots,X_m$ is a block-diagonal matrix with the $m$ matrices as diagonal submatrices.}
of a number of smaller matrices.

$s,t=1,\cdots,m$
$p,q=1,\cdots,r$
$i,j,k,l=1,\cdots,n$

For our purposes a slightly more general parametrization of stress is more
convenient. Suppose $Y_1,\cdots,Y_m$, with $m\leq np$, are linear independent
elements of configuration space $\mathbb{R}^{n\times p}$. We now parametrize the
configuration $X$ as a linear combination
\begin{equation}
X=\sum_{s=1}^m\alpha_s Y_s\label{eq-alpha}
\end{equation}
of the matrices $Y_s$ in the basis. We will study a version of smacof
parametrized using *coefficient space* $\mathbb{R}^m$, the space of all vectors
with the coefficients $\alpha$ of the linear combinations.

If $m=np$ then the alternative parametrization \eqref{eq-alpha} is merely a
linear change of coordinates. Stress from \eqref{eq-stressx} is the special case
in which each of the $np$ matrices $Y_s$ have a single element equal to one with
all other elements equal to zero. If $m<np$ the constraint can be used to impose
restrictions on the configuration (@bentler_weeks_78, @deleeuw_heiser_C_80).

The parametrization in \eqref{eq-alpha} makes the distances, and thus stress, a
function of $\alpha$. We will overload the symbols $\sigma$ and $d$, and define
stress as
\begin{equation}
\sigma(\alpha):=
\frac12\mathop{\sum}_{1\leq i<j\leq n} w_{ij}(\delta_{ij}-d_{ij}(\alpha))^2,
\label{eq-stressa}
\end{equation}

\sectionbreak

# Notation

First some convenient notation in configuration space, introduced in
@deleeuw_C_77. Vector $e_i$ has $n$ elements, with element $i$ equal to $+1$,
and all other elements zero. 
$$
A_{ij}:=(e_i-e_j)(e_i-e_j)'
$$
$(e_i-e_j)(e_i-e_j)'$, which
means elements $(i,i)$ and $(j,j)$ are equal to $+1$, while $(i,j)$ and $(j,i)$
are $-1$.

$$
\{H_{ij}\}_{st}:=\text{tr}\ Y_s'A_{ij}Y_t
$$
 Thus
\begin{equation}
d_{ij}^2(\alpha)=\text{tr}\ \alpha'A_{ij}\alpha,
\label{eq-anot}
\end{equation}

We also define
\begin{equation}
V:=\mathop{\sum}_{1\leq i<j\leq n}w_{ij}A_{ij},
\label{eq-vdef}
\end{equation}
and the matrix-valued function $B$ with
\begin{equation}
B(X):=\mathop{\sum}_{1\leq i<j\leq n}w_{ij}r_{ij}(X)A_{ij},
\label{eq-bdef}
\end{equation}
where
\begin{equation}
r_{ij}(X):=\begin{cases}\frac{\delta_{ij}}{d_{ij}(X)}&\text{ if }d_{ij}(X)>0,\\
0&\text{ if }d_{ij}(X)=0.\end{cases}\label{eq-rdef}
\end{equation}

If we assume, without loss of generality, that
\begin{equation}
\frac12\mathop{\sum}_{1\leq i<j\leq n}w_{ij}\delta_{ij}^2=1,
\label{eq-dnorm}
\end{equation}
then
\begin{equation}
\sigma(X)=1-\text{tr}\ X'B(X)X+\frac12\text{tr}\ X'VX.
\label{eq-S}
\end{equation}

$V$ is symmetric with non-positive off-diagonal elements. It is doubly-centered
(rows and columns add up to zero) and thus weakly diagonally dominant. It
follows that it is positive semi-definite (see @varga_62, section 1.5). Because
of irreducibility it has rank $n-1$, and the vectors in its null space are all
proportional to $e$, the vector with all elements equal to $+1$. The matrix
$B(X)$ is also symmetric, positive semi-definite, and doubly-centered for each
$X$. It may not be irreducible, because, for example, $B(0)=0$.

For the notation in coefficient space we need the square, symmetric, positive
semi-definite matrices $A_{ij}$ of order $m$ with elements
\begin{equation}
\{A_{ij}\}_{st}:=\text{tr}\ Y_s'A_{ij}Y_t.\label{eq-atildef}
\end{equation}
The squared distance equals
\begin{equation}
d_{ij}^2(\alpha):=\alpha'A_{ij}\alpha.\label{eq-dtildef}
\end{equation}
Note that we continue to overload symbols such as $A_{ij}$ and $d$. The same is
true for $V$ from \eqref{eq-vdef}, which we define in coefficient space by using
definition \eqref{eq-atildef}, and for $\tilde B(\alpha)$ from \eqref{eq-bdef}.
In coefficient space
\begin{equation}
\sigma(\alpha):=
1-\alpha'B(\alpha)\alpha+\frac12\alpha'V\alpha.\label{eq-stildef}
\end{equation}
and the MDS problem is to minimize \eqref{eq-stildef} over
$\alpha\in\mathbb{R}^m$.

\sectionbreak

# SMACOF

Define the *Guttman Transform* of configuration $X$ as
\begin{equation}
\Gamma(X):=V^+B(X)X
\label{eq-guttmanx}
\end{equation}
The *SMACOF algorithm* in configuration space is, with iteration counter
$k\geq 1$,
\begin{equation}
X^{(k)}=\Gamma(X^{(k-1)})=\Gamma^k(X^{0}),
\label{eq-smacofx}
\end{equation}
where $\Gamma^0=I$ and $\Gamma^k(X):=\Gamma(\Gamma^{k-1}(X))$.

In coefficient space the Guttman Transform is
\begin{equation}
\Gamma(\alpha):=V^+B(\alpha)\alpha
\label{eq-guttmana}
\end{equation}
The *SMACOF algorithm* in configuration space is
\begin{equation}
\alpha^{(k)}=\Gamma(\alpha^{(k-1)})=\Gamma^k(\alpha^{0}).
\label{eq-smacofx}
\end{equation}

@deleeuw_C_77 shows that the SMACOF iterations \eqref{eq-smacof} tend (in a
specific sense described later in this paper) to a fixed point of the Guttman
transform, i.e. to an $X$ with $\Gamma(X)=X$. If stress is differentiable at $X$
then the fixed points of the Guttman transform are stationary points, where the
derivative of stress vanishes. Also $V^+B(X)X=X$ shows that the columns of $X$
are eigenvectors of $V^+B(X)$, with eigenvalues equal to one.

The algorithm (without weights) was proposed by @guttman_68, by differentiating
stress and setting the derivatives equal to zero. This derivation ignores the
problem that stress is not differentiable if one or more distances are zero, and
it also does not say anything about convergence of the algorithm. In
@deleeuw_C_77 and @deleeuw_heiser_C_77 a new derivation of the algorithm was
given, using a majorization argument based on the Cauchy-Schwarz inequality.
This make it possible to avoid problems with differentiability, and it leads to
a simple convergence proof.

## Coefficient Space

Define the *Guttman Transform* of coefficients $\alpha$ as
\begin{equation}
\tilde\Gamma(\alpha):=\tilde V^+\tilde B(\alpha)\alpha
\label{eq-guttmana}
\end{equation}
The *SMACOF algorithm* is
\begin{equation}
\alpha^{(k+1)}=\tilde\Gamma(\alpha^{(k)})=\tilde\Gamma^k(\alpha^{0})
\label{eq-smacofa}
\end{equation}

\sectionbreak

# Global Convergence of SMACOF

Following @deleeuw_C_77 we also define
\begin{align}
\rho(X)&=\mathop{\sum}_{1\leq i<j\leq n} w_{ij}\delta_{ij}d_{ij}(X)=\text{tr}\ X'B(X)X,\\
\eta^2(X)&=\mathop{\sum}_{1\leq i<j\leq n} w_{ij}d_{ij}^2(X)=\text{tr}\ X'VX,
\end{align}
and
\begin{equation}
\lambda(X)=\frac{\rho(X)}{\eta(X)}.
\end{equation}
The main result in @deleeuw_C_77 is:

1. The SMACOF iterates $X^{(k)}$ are in the compact set $\eta(X)\leq 1$,
2. $\{\rho(X^{k})\}, \{\eta^(X^{k})\}$ and $\{\lambda(X^{k})\}$ are increasing
   sequences, with limits $\rho_\infty,\eta_\infty$ and $\lambda_\infty$, where
   $\eta_\infty=\lambda_\infty=\sqrt{\rho_\infty}$.
3. $\{\sigma(X^{k})\}$ is a decreasing sequence converging to
   $\sigma_\infty=1-\eta^2_\infty$.
4. At any accumulation point $X_\infty$ of $\{\sigma(X^{k})\}$ we have
   $\sigma(X_\infty)=\sigma_\infty$ and $X_\infty=\Gamma(X_\infty)$.
5. $\eta(X^{(k+1)}-X^{(k)})\rightarrow 0$ and thus either $\{\sigma(X^{k})\}$
   converges or the set of accumulation points of $\{\sigma(X^{k})\}$ is a
   continuum.

So, in words, the sequence of stress values converges monotonically. The
sequence of configurations is asymptotically regular and has one or more
accumulation points. Each accumulation point is a fixed point of the Guttman
transform. All accumulation points have the same function value, which is also
the limit of the stress sequence.

But it is important to emphasize that @deleeuw_C_77 did not show that the
configuration sequence $\{X^{(k)}\}$ actually a Cauchy sequence, and converges
to a configuration $X_\infty$. This is basically because of the rotational
invariance of the MDS problem. If $K$ is a rotation matrix, i.e. $K'K=KK'=I$,
then $d_{ij}(XK)=d_{ij}(X)$ for all $i$ and $j$, and thus
$\sigma(XK)=\sigma(X)$. If $X$ is a fixed point of the Guttman tranform, then so
is $XK$. Consequently there are no isolated fixed points, each fixed point is
part of a nonlinear continuum of fixed points.

It is also of interest that @deleeuw_A_84f showed that if $X$ is a local
minimizer of stress then $d_{ij}(X)>0$ for all $i$ and $j$ such that
$w_{ij}\delta_{ij}>0$. Thus, if weights and dissimilarities are positive, stress
is differentiable at a local minimum. @deleeuw_R_93c shows that stress has only
a single local maximum at $X=0$, and consequently the only possible stationary
points are local minima and saddle points. It is shown by @deleeuw_E_19g that
all fixed points of the Guttman transform which are not of full column rank are
saddle points. Thus if $X$ is a local minimizer of MDS, then adding $q$ zero
columns to $X$ produces a saddle point $[X\mid 0]$ for (p+q)MDS. At saddle
points there are always directions in which stress can be decreased, and
consequently the SMACOF algorithm with enough iterations will eventually get
close to a continuum of local minima.

\sectionbreak

# Local Convergence of SMACOF

@deleeuw_A_88b studies the local convergence of SMACOF, i.e. the *rate of
convergence* of the SMACOF iterations. There is, however, a problem not
adequately addressed in that article, which has quite a bit of hand-waiving, and
some incorrect statements. If there is no convergence to a single configuration
then the usual rate of convergence is not defined either.

We know the sequence $\{\eta(X^{(k+1)}-X^{(k)})\}$ tends to zero. We can measure
its rate of convergence by the *ratio factor*
\begin{equation}
q_k=\frac{\eta(X^{(k+1)}-X^{(k)})}{\eta(X^{(k)}-X^{(k-1)})},
\end{equation}
and the *root factor*
\begin{equation}
r_k=\eta(X^{(k+1)}-X^{(k)})^{1/k}.
\end{equation}
There is no guarantee that either $\{q_k\}$ or $\{r_k\}$ converges, but
\begin{equation}
\liminf_{k\rightarrow\infty} q_k\leq\liminf_{k\rightarrow\infty} r_k\leq\limsup_{k\rightarrow\infty} r_k\leq\limsup_{k\rightarrow\infty} q_k
\end{equation}
Thus if $\lim_{k\rightarrow\infty} q_k$ exists then
$\lim_{k\rightarrow\infty} r_k$ exist and the two limit values are equal. See
@rudin_76, p. 68, theorem 3.37.

The rate of convergence of the SMACOF iterations at a fixed point $X$ is
computed in @deleeuw_A_88b as the spectral norm
$\kappa(X)=\|\mathcal{D}\Gamma(X)\|_\infty$, i.e. as the modulus of the largest
eigenvalue of the derivative of the Guttman transform. There are good reasons to
compute this quantity. From Ostrovski's Theorem (@ostrowski_73, theorem 22.1, p.
151 , @ortega_rheinboldt_70, theorem 10.1.3, p. 300) we know that if
$\kappa(X)<1$, then $X$ is a *point of attraction* of the SMACOF iteration. If
we start close enough then the iterations will converge to $X$. We also know
(@ostrowski_73, theorem 22.2, p. 152) that if $\kappa(X)>1$ then $X$ is a *point
of repulsion*, which means the iterations will diverge when started in a certain
solid angle with $X$ at its apex.

If we define, with @ortega_rheinboldt_70, chapter 9, for a sequence
$\{X^{(k)}\}$ converging to $X$, the ratio and root convergence factors
\begin{equation}
Q_1(\{X^{(k)}\})=\limsup_{k\rightarrow\infty}\frac{\eta(X^{(k+1)}-X)}{\eta(X^{(k)}-X)},
\end{equation}
and
\begin{equation}
R_1(\{X^{(k)}\})=\limsup_{k\rightarrow\infty}\ \eta(X^{(k)}-X)^{1/k},
\end{equation}
then $\kappa(X)<1$ implies $R_1(X^{(k)})=\kappa(X)$ (@ortega_rheinboldt_70,
theorem 10.1.4, p. 301). Alo $R_1(X^{(k)})$ is independent of the norm we have
chosen, which is $\eta$ in the SMACOF case (@ortega_rheinboldt_70, theorem
9.9.2, p. 288). Because $Q_1(X^{(k)})$ depends on the norm we can only assert
that $Q_1(X^{(k)})\geq R_1(X^{(k)})=\kappa(X)$, although it is guaranteed in the
SMACOF case that some norm exists for which there is equality
(@ortega_rheinboldt_70, Notes and Remarks NR 10.1-5, p. 306).

There is a problem, however, with applying the Ostrowski and Ortega-Rheinboldt
results in our case. Because of the invariance of distances under rotation we
have $\kappa(X)=1$, no matter what $X$ is. And, basically for the same reason,
we have not shown that the SMACOF iterations converge to a fixed point at all.
@deleeuw_A_88b on page 171 acknowledges the problem. In fact, he proposes a way
around the problem in the proof of theorem 3 on page 175, but the proof is not
very convincing, although convincing enough to get past the reviewers. I will
try to be more specific and precise.

Observe that
\begin{equation}
\mathcal{D}\Gamma(X)=V^+\mathcal{D}^2\rho(X),
(\#eq:dgamma)
\end{equation}
and
\begin{equation}
\mathcal{D}^2\sigma(X)=V-\mathcal{D}^2\rho(X)=V(I-\mathcal{D}\Gamma(X)).
(\#eq:d2sigma)
\end{equation}
Because $\rho$ is convex $\mathcal{D}^2\rho(X)$ is symmetric and positive
semi-definite. The eigenvalues of $\mathcal{D}\Gamma(X)$ are the generalized
eigenvalues of the positive semi-definite matrix pair
$(\mathcal{D}^2\rho(X),V)$.

An explicit formula for the derivatives of the Guttman transform, or
equivalently for the second derivatives of stress, was first given by
@deleeuw_A_88b. We write $\mathcal{D}\Gamma(X)[Y]$ for the derivative evaluated
at $Y$. Partial derivatives can be obtained by choosing $Y=e_ie_s'$. Of course
we assume here that $d_{ij}(X)>0$ for all $i<j$ with $w_{ij}\delta_{ij}>0$.
\begin{equation}
\mathcal{D}\Gamma(X)[Y]=V^+\{B(X)Y-H(X,Y)X\},
(\#eq:dgammamat)
\end{equation}
with
\begin{equation}
H(X,Y)=\mathop{\sum}_{1\leq i<j\leq n}w_{ij}\frac{\delta_{ij}}{d_{ij}(X)}\frac{\text{tr}\ X'A_{ij}Y}{d_{ij}^2(X)}A_{ij}.
(\#eq:hdef)
\end{equation}
These formulas, together with equations \@ref(eq:dgamma) and \@ref(eq:d2sigma),
prove the main results in @deleeuw_A_88b.

1. All eigenvalues of $\mathcal{D}\Gamma(X)$ are non-negative.
2. **(Homogeneity)** $X$ is an eigenvector of $\mathcal{D}\Gamma(X)$ with
   eigenvalue zero.
3. **(Translation)** $\mathcal{D}\Gamma(X)$ has at least $p$ additional
   eigenvalues equal to zero, corresponding with eigenvectors of the form
   $ee_s'$, where $e$ has all elements equal to one and $e_s$ has one of its
   elements equal to one and the others equal to zero.
4. **(Rotation)** If $X$ is a fixed point of the Guttman transform then
   $\mathcal{D}\Gamma(X)$ has at least $\frac12 p(p-1)$ eigenvalues equal to
   one, corresponding with eigenvectors of the form $XA$, with $A$
   anti-symmetric.
5. At a local minimum point $X$ of stress all eigenvalues of
   $\mathcal{D}\Gamma(X)$ are less than or equal to one. At a strict local
   minimum they are strictly less than one.
6. At a saddle point $X$ of stress the largest eigenvalue of
   $\mathcal{D}\Gamma(X)$ is larger than one.

# Modified SMACOF

To avoid the problems caused by a continuum of fixed points @deleeuw_A_88b
proposes to rotate each SMACOF iterate to a unique position, for example to
principal components. For any $X$ with singular value decomposition
$X=K\Lambda L'$ we define $\Pi(X)=K\Lambda=XL.$ Note that if one or more
singular values are equal then $\Pi$ is not unique and becomes a point-to-set
map. We can rotate within each of the spaces corresponding to multiple singular
values, and thus we have not completely eliminated indeterminacy due to
rotation. In the sequel we simply assume that the $p$ singular values of $X$ are
different.

Now consider the iterations
\begin{equation}
\tilde X^{(k+1)}=\Pi(\Gamma(\tilde X^{(k)}))={(\Pi\Gamma)}^k(\tilde X^{0}).
\end{equation}
Let's call this modified SMACOF, or mSMACOF for short.

It looks initialy as if mSMACOF require a great deal of extra computation, but
this is actually not the case. If $M$ is a rotation matrix we have
$\Pi(XM)=\Pi(X)$ and $\Gamma(XM)=\Gamma(X)M$. This implies that
$\Pi\Gamma\Pi(X)=\Pi\Gamma(X)$ for all $X$. As a result the sequences
$\{X^{(k)}\}$ and $\{\tilde X^{(k)}\}$ have a simple relationship. If we start
the $\{\tilde X^{(k)}\}$ sequence with $X^{0}$ then for each $k$ we have
$\sigma(\tilde X^{(k)})=\sigma(X^{(k)})$ and $\tilde X^{(k)}=\Pi(X^{(k)})$. Thus
the two sequences of stress values are exactly the same for SMACOF and mSMACOF
and the two configurations, and consequnetly also the accumulation points of the
two sequences of configurations, just differ by a rotation. To get
$\tilde X^{(k)}$ we can just compute the SMACOF iterate $X^{(k)}$ and rotate it
to principal components.

The derivative of $\Pi\Gamma$ is, by the chain rule,
\begin{equation}
\mathcal{D}\Pi\Gamma(X)=\mathcal{D}\Pi(\Gamma(X))\mathcal{D}\Gamma(X).
\end{equation}
After some computation we find
\begin{equation}
\mathcal{D}\Pi(X)[Y]=YL+K\Lambda M,
\end{equation}
where $X=K\Lambda L'$ and $M$ is the anti-symmetric matrix with off-diagonal
elements
\begin{equation}
m_{ij}=-\frac{\lambda_iu_{ij}+\lambda_ju_{ji}}{\lambda_i^2-\lambda_j^2}
\end{equation}
with $U=K'YL$.

If $X$ satisfies $\Pi(X)=X$ then $L=I$ and thus $\mathcal{D}\Pi(X)[Y]=Y+XM$ and
$U=K'Y$. We proceed to compute the eigenvectors and eigenvalues of the Jacobian
at an $X$ with $\Pi(X)=X$.

1. If $Y=XA$ with $A$ antisymmetric then $M=-L'AL$ and $\mathcal{D}\Pi(X)[Y]=0$.
   This corresponds with $\frac12 p(p-1)$ zero eigenvalues
2. If $Y=K\Lambda^{-1}A$ with $A$ antisymmetric then $M=0$. This gives
   $\frac12 p(p-1)$ eigenvectors with eigenvalue one.
3. If $Y=K\Lambda^{-1}D$ wth $D$ diagonal then $M=0$. This gives another $p$
   eigenvectors with eigenvalue one.
4. If $Y=K_\perp S$ with $K_\perp$ a basis for the orthogonal complement of $X$
   then $M=0$ and we have another $p(n-p)$ eigenvalues equal to one.

This is a complete set of eigenvectors. It turns out all eigenvalues are equal
to one, except for $\frac12 p(p-1)$ zero eigenvalues. It follows that fixed
points of $\Pi\Gamma$ are of the form $\tilde X=XL$, with $X$ a fixed point of
$\Gamma$ and $L$ the rotation of $X$ to principal components). Thus
$\Gamma(\tilde X)=\Gamma(X)L=XL=\tilde X$, and $\tilde X$ is a fixed point of
both $\Gamma$ and $\Pi\Gamma$. The eigenvalues of
$\mathcal{D}\Pi\Gamma(\tilde X)$ are the same as the eigenvalues of
$\mathcal{D}\Gamma(X)$, except for the $\frac12 p(p-1)$ eigenvalues equal to
one, corresponding with $XA$ for antisymmetric A, which are replaced by zero
eigenvalues. The Ostrowski and Ortega-Rheinboldt results now apply directly to
$\Pi\Gamma$. If there are no zero distances, if the singular values of $X$ are
different, and if the largest eigenvalue of $\mathcal{D}\Pi\Gamma(X)$ is less
than one, then $X$ is a point of attraction, the SMACOF iterations converge to
$X$ if started sufficiently close to it (in the *attraction ball* of $X$), and
$R_1(X)$ is equal to the largest eigenvalue.

# Software

The appendix has an ad-hoc version of the `smacof` function. Its arguments are

```{r}
#| label: smacof_args

args(smacof)
```
The function expects both dissimilarities and weights to be of class `dist`. The
default initial configuration uses classical MDS (@torgerson_58). We iterate
until either the distance $\eta(X^{(k)}-X^{(k-1)})$ between successive
configurations is less than `eps`, which has the ridiculously small default
value of `r 1e-15`, or until the maximum number of iterations `itmax` is
reached, which defaults to the ridiculously large default value of 10000. If
`pca` is TRUE all iterates are rotated to principal components. If `verbose` is
TRUE we print, for each iteration, the iteration number $k$, the stress
$\sigma(X^{(k)})$, the change $\eta(X^{(k)}-X^{(k-1)})$, and the root and ratio
convergence factors.

The second file `derivative.R` computes the Jacobian of the iteration functions
at the limit (i.e. at the point where SMACOF stops). There are actually six
functions in the file. The three functions `dGammaA, dPiA` and `dPiGammaA` use
the analytical expressions we have derived earlier for
$\mathcal{D}\Gamma(X), \mathcal{D}\Pi(X)$ and $\mathcal{D}\Pi\Gamma(X).$ The
corresponding functions `dGammaN, dPiN` and `dPiGammaN` numerically compute the
Jacobian using the `jacobian` function from the `numDeriv` package
(@gilbert_varadhan_19). They are basically used to check the formulas.

# Examples

## Ekman

Time for a numerical example. We will use the color similarity data of
@ekman_54, taken from the SMACOF package (@deleeuw_mair_A_09c). We transform
Ekman's similarities $s_{ij}$ to dissimilarities $\delta_{ij}$ using
$\delta_{ij}=(1-s_{ij})^3$, and we use unit weights in $W$.

```{r}
#| label: ekman_data
#| echo: false

data (ekman, package = "smacof")
delta <- (1 - ekman) ^ 3
weights <- rep (1, length (ekman))
attributes (weights) <- attributes (ekman)
delta <- delta / sqrt (sum (weights * delta ^ 2) / 2)
```

```{r}
#| label: ekman_smacof
#| echo: false

h <- smacof (delta, weights, eps = 1e-15)
```
SMACOF requires `r h$itel` iterations for convergence. The minimum of stress is
`r h$s`, the root factor is equal to `r h$r` and the ratio factor is `r h$q`.
The eigenvalues of $\mathcal{D}\Gamma(X)$ are

```{r}
#| label: ekman_eval_h
#| echo: false

hDerA <- eigen (dGammaA (h$x, delta, weights))$values
for (i in 1:28) {
  cat (
    formatC(Mod(hDerA[i]), digits = 15, format = "f"),
    "\n"
  )
}
```
As an aside, the Ekman example (with this particular transformation of the
similarities) is special, because the two-dimensional solution is actually the
global minimum over all configurations (i.e. the global mnimum of MDS for all
$1\leq p\leq n$). As shown in @deleeuw_U_14b this follows from the fact that the
two unit eigenvalues of $V^+B(X)$ are actually its two largest eigenvalues, and
consequently $X$ is the solution of the convex full-dimensional nMDS relaxation
of the MDS problem (@deleeuw_groenen_mair_E_16e). The eigenvalues of $V^+B(X)$
are

```{r}
#| label: ekman_eval_b
#| echo: false

for (e in eigen(h$b)$values) {
  cat (formatC(e, digits = 15, format = "f"), "\n")
}
```

```{r}
#| label: ekman_smacof_mod
#| echo: false

h <- smacof (delta, weights, pca = TRUE, eps = 1e-15)
```
We now set `pca=TRUE` and run mSMACOF. It requires `r h$itel` iterations for
convergence. The minimum of stress is `r h$s`, and the root factor is equal to
`r h$r` and the ratio factor is `r h$q`. The eigenvalues of
$\mathcal{D}\Pi\Gamma(X)$ at the fixed point $X$ are

```{r}
#| label: ekman_eval_h_mod
#| echo: false

hDerA <- eigen (dPiGammaA (h$x, delta, weights))$values
for (i in 1:28) {
  cat (
    formatC(Mod(hDerA[i]), digits = 15, format = "f"),
    "\n"
  )
}
```
They are equal to the eigenvalues of $\mathcal{D}\Gamma(X)$ and the root
convergence factor $R_1(\{X^{(k)}\})$ is `r max(Mod(hDerA))`.

## De Gruijter

The second example are dissimilarties between nine Dutch political parties,
collected a long time ago by @degruijter_67.

```{r}
#| label: poldist_data
#| echo: false

poldist <-
structure(c(5.63, 5.27, 4.6, 4.8, 7.54, 6.73, 7.18, 6.17, 6.72, 
5.64, 6.22, 5.12, 4.59, 7.22, 5.47, 5.46, 4.97, 8.13, 7.55, 6.9, 
4.67, 3.2, 7.84, 6.73, 7.28, 6.13, 7.8, 7.08, 6.96, 6.04, 4.08, 
6.34, 7.42, 6.88, 6.36, 7.36), Labels = c("KVP", "PvdA", "VVD", 
"ARP", "CHU", "CPN", "PSP", "BP", "D66"), Size = 9L, call = quote(as.dist.default(m = polpar)), class = "dist", Diag = FALSE, Upper = FALSE)
poldist <- poldist - 3
print (poldist, digits = 3)
weights <- rep (1, length (poldist))
attributes (weights) <- attributes (poldist)
poldist <- poldist / sqrt (sum (weights * poldist ^ 2) / 2)
```
We analyze the data with SMACOF using three dimensions.

```{r}
#| label: poldist_smacof
#| echo: false

h <- smacof (poldist, weights, p = 3, eps = 1e-15)
```
SMACOF requires `r h$itel` iterations for convergence. The minimum of stress is
`r h$s`, the root factor is equal to `r h$r`, and the ratio factor is `r h$q`.
The eigenvalues of $\mathcal{D}\Gamma(X)$ are

```{r}
#| label: poldist_eval_h
#| echo: false

hDerA <- eigen (dGammaA (h$x, poldist, weights))$values
for (i in 1:length(hDerA)) {
  cat (
    formatC(Mod(hDerA[i]), digits = 15, format = "f"),
     "\n"
  )
}
```
In this case there is no guarantee that we have the three-dimensional global
minimum. The unit eigenvalues of the matrix $V^+B(X)$ are not the three largest
ones.

```{r}
#| label: poldist_eval_b
#| echo: false

for (e in eigen(h$b)$values) {
  cat (formatC(e, digits = 15, format = "f"), "\n")
}
```

```{r}
#| label: poldist_msmacof
#| echo: false

h <- smacof (poldist, weights, p = 3, pca = TRUE, eps = 1e-15)
```
mSMACOF in three dimensions requires `r h$itel` iterations for convergence. The
minimum of stress is `r h$s`, the root factor is equal to `r h$r`, and the ratio
factor is `r h$q`. The eigenvalues of $\mathcal{D}\Pi\Gamma(X)$ are

```{r}
#| label: poldist_eval_mh
#| echo: false

hDerA <- eigen (dPiGammaA (h$x, poldist, weights))$values
for (i in 1:length(hDerA)) {
  cat (
    formatC(Mod(hDerA[i]), digits = 15, format = "f"),
     "\n"
  )
}
```
Again, as expected, the eigenvalues for SMACOF and mSMACOF are the same and the
root convergence factor is the spectral norm of the mSMACOF derivative, which is
`r max(Mod(hDerA))`.

# Conclusion

From a practical point of view there is no need to ever use modified SMACOF. We
only use the Guttman transform, and compute the largest eigenvalue of its
derivative at the solution $X$, ignoring the $\frac12 p(p-1)$ trivial unit
eigenvalues. That largest eigenvalue, if it is strictly smaller than one, is the
root convergence factor. In that case SMACOF can be said to converge to
$\Pi(X)$, which is the SMACOF solution rotated to principal components.

\sectionbreak 

# References
